---
title: "Massbank2SIRIUS"
author: "Ida"
date: "2023-01-10"
output: html_document
---
Code for calculating the fingerprint features with SIRIUS+CSI:FingerID

```{r}
library(data.table)
library(dplyr)
library(varhandle)
library(stringr)
library(rJava)
library(webchem)
library(enviPat)
```

Reading in the real-life test set (generated in `Split_data_real_life_test_set.ipynb`) and adding the needed information about the experimental conditions used while measuring the MS2 spectra.
```{r}
validation_set <- as.data.frame(fread('validation_set_final.tsv', sep='\t', header=T))

full_name_creator <- function(file_name) {
  return(paste0('./MassBank-data-2022.06a/', strsplit(file_name, '-')[[1]][2], '/', file_name))
}

files_list <- unlist(lapply(validation_set$file_name, full_name_creator))
validation_set$full_path <- files_list

validation_set$ms_type <- NA
validation_set$precursor_MZ <- NA
validation_set$formula <- NA
validation_set$precursor_type <- NA
validation_set$instrument <- NA

counter = 1
for (file in files_list) {
    file_connection <- file(file)
    raw_data <- readLines(file_connection)
    close(file_connection)
    ms_type <- substring(grep('AC$MASS_SPECTROMETRY: MS_TYPE', raw_data, value=T, fixed=T), 31)
    validation_set[counter, 'ms_type'] <- ms_type
    if (ms_type == 'MS2') {
      precursor_MZ <- substring(grep('MS$FOCUSED_ION: PRECURSOR_M/Z', raw_data, value=T, fixed=T), 31)
      validation_set[counter, 'precursor_MZ'] <- ifelse(length(precursor_MZ) == 0, '', precursor_MZ)
      formula <- substring(grep('CH$FORMULA:', raw_data, value = T, fixed=T), 13)
      if (substr(formula, 1, 1) == '[' & substr(formula,nchar(formula)-1,nchar(formula)) == ']+') {
        formula <-  gsub('[|[+]|]|', '', formula)
      }
      validation_set[counter, 'formula'] <- formula
      precursor_type <- substring(grep('MS$FOCUSED_ION: PRECURSOR_TYPE', raw_data, value=T, fixed=T), 32)
      if (length(precursor_type) == 0) {
        precursor_type <- substring(grep('MS$FOCUSED_ION: ION_TYPE', raw_data, value=T, fixed=T), 26)
        if (length(precursor_type) == 0) {
         is_neg <- grepl('NEGATIVE', substring(grep('AC$MASS_SPECTROMETRY: ION_MODE', raw_data, value=T, fixed=T), 0))
         is_pos <- grepl('POSITIVE', substring(grep('AC$MASS_SPECTROMETRY: ION_MODE', raw_data, value=T, fixed=T), 0))
         if (is_neg == T) {
           precursor_type <- '[M-H]-'
         }
         if (is_pos == T) {
           precursor_type <- '[M+H]+'
         }
        }
      }
      validation_set[counter, 'precursor_type'] <- precursor_type
      instrument <- substring(grep('AC$INSTRUMENT:', raw_data, value=T, fixed=T), 16)
      validation_set[counter, 'instrument'] <- ifelse(length(instrument) == 0, '', instrument)
    }
    counter = counter + 1
}

validation_set <- validation_set %>% filter(ms_type == 'MS2')
validation_set <- validation_set %>% group_by(SMILES) %>% mutate(precursor_MZ = max(precursor_MZ))
validation_set <- validation_set %>% group_by(SMILES, precursor_type) %>%
                  dplyr::mutate(unique_name = paste0('validation_', cur_group_id()))
validation_set[validation_set$precursor_type == '[M+H-NH3]+', 'precursor_type'] <- '[M+NH4]+'

write.table(validation_set, 'validation_set_before_final_filter_with_groupIDs.tsv', sep='\t', col.names=F, row.names=F, quote=F)

validation_set$instrument_SIRIUS <- ifelse(grepl('QTOF', validation_set$instrument_type), 'qtof',
                                           ifelse(grepl('orbitrap', tolower(validation_set$instrument)) | 
                                                  grepl('Q Exactive Thermo Fisher Scientific', validation_set$instrument) |
                                                  grepl('Q-Exactive HF', validation_set$instrument),
                                                  'orbitrap', ''))

val1 <- validation_set %>% filter(precursor_type == '[M+H]+')
SMILES1 <- unique(val1$SMILES)
val2 <- validation_set %>% filter(precursor_type == '[M+NH4]+' & !SMILES %in% SMILES1)
val <- rbind(val1, val2)
SMILES2 <- unique(val$SMILES)
val3 <- validation_set %>% filter(precursor_type == '[M+Na]+' & !SMILES %in% SMILES2)
val <- rbind(val, val3)
SMILES3 <- unique(val$SMILES)
val4 <- validation_set %>% filter(precursor_type == '[M-H]-' & !SMILES %in% SMILES3)
val <- rbind(val, val4)
SMILES4 <- unique(val$SMILES)
val5 <- validation_set %>% filter(precursor_type == '[M]+' & !SMILES %in% SMILES4)
val <- rbind(val, val5)
SMILES5 <- unique(val$SMILES)
validation_set_final_groups <- val
write.table(validation_set_final_groups, 'validation_set_with_groupIDs.tsv', sep='\t', col.names=F, row.names=F, quote=F)

validation_set_fg_orbitrap <- validation_set_final_groups %>% filter(instrument_SIRIUS == 'orbitrap')
SMILES_orbitrap <- unique(validation_set_fg_orbitrap$SMILES)
unique_names_orbitrap <- unique(validation_set_fg_orbitrap$unique_name)
write.table(validation_set_fg_orbitrap, 'validation_set_orbitrap.tsv', sep='\t', col.names=F, row.names=F, quote=F)

validation_set_fg_qtof <- validation_set_final_groups %>% filter(instrument_SIRIUS == 'qtof' & !SMILES %in% SMILES_orbitrap)
SMILES_qtof <- unique(validation_set_fg_qtof$SMILES)
unique_names_qtof <- unique(validation_set_fg_qtof$unique_name)
write.table(validation_set_fg_qtof, 'validation_set_qtof.tsv', sep='\t', col.names=F, row.names=F, quote=F)

problematic <- validation_set_final_groups %>% filter(!unique_name %in% unique_names_orbitrap & !unique_name %in% unique_names_qtof)

write.table(rbind(validation_set_fg_orbitrap, validation_set_fg_qtof), 'validation_set_SIRIUS.tsv', sep='\t', col.names=F, row.names=F, quote=F)

```

Calculating the MS1 data and generating the `.ms`files for SIRIUS
```{r}
ms1_generator <- function(formula, charge){
  data(isotopes)
  pattern <- isopattern(isotopes,
                        formula,
                        threshold=0.1,
                        plotit=FALSE,
                        charge = charge,
                        emass = 0.00054858,
                        algo=1)
  isotope <- as.data.frame(pattern[[1]]) %>% 
             dplyr::select(`m/z`,abundance)
  names(isotope)[names(isotope) == 'm/z'] <- 'V1'
  names(isotope)[names(isotope) == 'abundance'] <- 'V2'
  return(isotope)
}

file_for_SIRIUS <- function(compound_files, compound, folder, precursor_type, precursor_MZ, formula) {
  ms2_data <- tibble()
  empty_row <- c('','')
  for (file in compound_files) {
    file_connection <- file(file)
    raw_data <- readLines(file_connection)
    close(file_connection)

    collision_energy <- substring(grep('AC$MASS_SPECTROMETRY: COLLISION_ENERGY', raw_data, value=T, fixed=T), 40)
    if (length(collision_energy) == 0) {
      print(file)
    } else {
    collision_energy <- sub('\\ .*', '', collision_energy)
    if (check.numeric(collision_energy) == F){
      collision_energy = '99'
    }
    cas <- substring(grep('CH$LINK: CAS', raw_data, value=T, fixed=T), 14)
    if (identical(cas, character(0))){
      cas = 'CASmissing'
    }
    DTX <- substring(grep('CH$LINK: COMPTOX', raw_data, value=T, fixed=T), 18)
    if (identical(DTX, character(0))){
      DTX = 'DTXmissing'
    }
    retention_time <- substring(grep('AC$CHROMATOGRAPHY: RETENTION_TIME', raw_data, value=T, fixed=T), 35)
    retention_time <- gsub(' ', '', retention_time, fixed=T)
    if (identical(retention_time, character(0))){
      ret_time = 'RTmissing'
    }
    peak_start <- grep('PK$PEAK:', raw_data, fixed=T) + 1
    peak_end <- tail(grep('//', raw_data, fixed=T), 1)

    if(peak_start < peak_end){
      splitted <- strsplit(raw_data[peak_start:(peak_end - 1)], ' ')
      peaks <- matrix(nrow = peak_end - peak_start, ncol = 3)
      for (i in 1:length(splitted)) {
        splitted[[i]] <- splitted[[i]][which(splitted[[i]] != '')]
        peaks[i,] <- splitted[[i]]
      }
      peaks <- as.data.frame(peaks, stringsAsFactors = FALSE)
      peaks[] <- lapply(peaks, type.convert)

      peaks_col <- c('>collision', collision_energy)
      peaks <- peaks %>% dplyr::select(-V3)
      ms2_tabel <- rbind(peaks_col, peaks, empty_row)
      ms2_data <- ms2_data %>% bind_rows(ms2_tabel)
    }
    }
  }
  
  compound_row <- c('>compound', compound)
  formula_row <- c('>formula', formula)
  parent_row <- c('>parentmass', precursor_MZ)
  ionisation_row <- c('>ionization', precursor_type)
  header_table <- rbind(compound_row, formula_row, parent_row, ionisation_row)
  if (grepl(']-', precursor_type, fixed=T)) {
    if (grepl('H', formula, fixed=T)){
      splitted_formula <- strsplit(formula, "(?=[A-Za-z])(?<=[0-9])|(?=[0-9])(?<=[A-Za-z])", perl=T)[[1]]
      no_Hs <- as.numeric(splitted_formula[which(splitted_formula == 'H') + 1])
      new_no_Hs <- no_Hs - 1
      splitted_formula[which(splitted_formula == 'H') + 1] <- new_no_Hs
      formula_ms1 <- paste0(splitted_formula, collapse='')
    } else {
      formula_ms1 <- formula
    }
    ms1_data <- ms1_generator(formula_ms1, -1)
  } else {
    formula_ms1 <- paste0(formula, strsplit(precursor_type, '[+]|]')[[1]][2])
    ms1_data <- ms1_generator(formula_ms1, 1)
  }
  ms1_name <- c(">ms1peaks","")
  
  SIRIUS_file <- rbind(header_table, empty_row, ms2_data, ms1_name, ms1_data)
  write.table(SIRIUS_file, paste0(folder, '/', compound, '.ms'), sep=' ', col.names=F, row.names=F, quote=F)
}

```


```{r}
# Orbitrap
compound_names_orbitrap <- unique(validation_set_fg_orbitrap$unique_name)
for (compound in compound_names_orbitrap) {
  file_for_SIRIUS((validation_set_fg_orbitrap %>% dplyr::filter(unique_name == compound))$full_path, 
                  compound, 
                  'SIRIUS_input/orbitrap',
                  (validation_set_fg_orbitrap %>% dplyr::filter(unique_name == compound))$precursor_type[1],
                  (validation_set_fg_orbitrap %>% dplyr::filter(unique_name == compound))$precursor_MZ[1],
                  (validation_set_fg_orbitrap %>% dplyr::filter(unique_name == compound))$formula[1])
}

# QTOF
compound_names_qtof <- unique(validation_set_fg_qtof$unique_name)
for (compound in compound_names_qtof) {
  file_for_SIRIUS((validation_set_fg_qtof %>% dplyr::filter(unique_name == compound))$full_path, 
                  compound, 
                  'SIRIUS_input/qtof',
                  (validation_set_fg_qtof %>% dplyr::filter(unique_name == compound))$precursor_type[1],
                  (validation_set_fg_qtof %>% dplyr::filter(unique_name == compound))$precursor_MZ[1],
                  (validation_set_fg_qtof %>% dplyr::filter(unique_name == compound))$formula[1])
}

```

Function for unpacking the data from SIRIUS5
```{r}
unzip_SIRIUS5 <- function(SIRIUS_data_folder) {
  subfolder_fp_zip <- dir(SIRIUS_data_folder, all.files=T, recursive=T, pattern='fingerprints')

  for (zipped_folder in subfolder_fp_zip){
    outfolder <- str_split(zipped_folder, '/')[[1]][1]
    outDir <- paste(SIRIUS_data_folder, outfolder, 'fingerprints1', sep='/')
    zipFile <- paste(SIRIUS_data_folder, zipped_folder, sep='/')
    unzip(zipFile, exdir = outDir)
  }

  subfolder_scores_zip <- dir(SIRIUS_data_folder, all.files=T, recursive=T, pattern='scores')

  for (zipped_folder in subfolder_scores_zip) {
    outfolder <- str_split(zipped_folder, '/')[[1]][1]
    outDir <- paste(SIRIUS_data_folder, outfolder, 'scores1', sep='/')
    zipFile <- paste(SIRIUS_data_folder, zipped_folder, sep='/')
    unzip(zipFile, exdir = outDir)
  }
    
  subfolder_fingerid_zip <- dir(SIRIUS_data_folder, all.files=T, recursive=T, pattern='fingerid')

  for (zipped_folder in subfolder_scores_zip) {
    outfolder <- str_split(zipped_folder, '/')[[1]][1]
    outDir <- paste(SIRIUS_data_folder, outfolder, 'fingerid1', sep='/')
    zipFile <- paste(SIRIUS_data_folder, zipped_folder, sep='/')
    unzip(zipFile, exdir = outDir)
  }
}
```

Function for calculating the fingerprint features with SIRIUS5
```{r}
MS2fp_SIRIUS <- function(MS_folder, SIRIUS_folder, output_folder, instrument, files_again = '') {
  setwd(MS_folder)
  filenames = ifelse(files_again == '', dir(pattern='*.ms'), files_again)
  for (file in filenames) {
    print(file)
    setwd(SIRIUS_folder)
    system(paste("cd ", SIRIUS_folder))
    command = str_glue(" .\\sirius ",
                       "-i ", str_glue('"', MS_folder, "/", file, '"', sep = ""),
                       " -o ", output_folder,
                       " formula",
                       #" formula -c 10",
                       " -p ", instrument,
                       " --ppm-max 5",
                       " --ppm-max-ms2 5",
                       " -E CH -e ONP[8]B[11]Si[9]S[12]Cl[18]Se[2]Br[10]FI[6]K[1]Na[1]As[2]",
                       " -d ALL",
                       " fingerprint", 
                       " structure",
                       " write-summaries",
                       sep =" ")
    system(command)
    tryCatch(
      expr = withTimeout(
        {
          javaOutput = system(command, intern = TRUE)
        },
        timeout = 300
      )
      ,
      error = function(e) {
        return(tibble())
      }
    )
  }
}
```

Calculating the fingerprint features with SIRIUS5
```{r}
MS_folder_orbitrap <- './SIRIUS_input/orbitrap'
MS_folder_qtof <- './SIRIUS_input/qtof'
SIRIUS_folder <- 'C:/Program Files/sirius'
output_folder <- './SIRIUS_output'

#MS2fp_SIRIUS(MS_folder_orbitrap, SIRIUS_folder, output_folder, 'orbitrap')
#MS2fp_SIRIUS(MS_folder_qtof, SIRIUS_folder, output_folder, 'qtof')
```

Checking which compounds do not have calculated fingerprint features
```{r}
validation_set_SIRIUS <- fread('validation_set_SIRIUS.tsv', sep='\t', header=F)
validation_set_SIRIUS_compounds <- (validation_set_SIRIUS %>% group_by(V23))[,c(1:13, 21, 23)]
validation_set_SIRIUS_compounds <- validation_set_SIRIUS_compounds[!duplicated(validation_set_SIRIUS_compounds), ]
colnames(validation_set_SIRIUS_compounds) <- c('SMILES', 'nr.ahr', 'nr.ar.lbd', 'nr.ar', 'nr.aromatase','nr.er.lbd', 'nr.er', 'nr.ppar.gamma', 'sr.are', 'sr.atad5', 'sr.hse', 'sr.mmp', 'sr.p53', 'ionization', 'validation_ID')

csi_fingerid <- fread(paste0(output_folder, '/csi_fingerid.tsv'), sep='\t', header=T)
csi_fingerid_neg <- fread(paste0(output_folder, '/csi_fingerid_neg.tsv'), sep='\t', header=T)

error_files <- c()
for (i in 1:nrow(validation_set_SIRIUS_compounds)) {
  folders <- list.files(output_folder, pattern=paste0(validation_set_SIRIUS_compounds$validation_ID[i], '_'), full.names=T)
  length_folders <- length(folders)
  folder_name = ''
  for (folder in folders) {
    if (length(list.files(paste0(folder, '/fingerprints'), pattern='.fpt', full.names=T)) == 1) {
      folder_name = list.files(paste0(folder, '/fingerprints'), pattern='.fpt', full.names=T)
    }
    if (length(list.files(paste0(folder, '/fingerprints1'), pattern='.fpt', full.names=T)) == 1) {
      folder_name = list.files(paste0(folder, '/fingerprints1'), pattern='.fpt', full.names=T)
    }
  }
  
  tryCatch(
        {
          data <- fread(folder_name)
          colnames(data) <- validation_set_SIRIUS_compounds$validation_ID[i]
          if (nrow(data) == nrow(csi_fingerid)) {
            csi_fingerid <<- cbind(csi_fingerid, data)
          }
          if (nrow(data) == nrow(csi_fingerid_neg)) {
            csi_fingerid_neg <<- cbind(csi_fingerid_neg, data)
          }
        },
        error=function(e) {
            error_files <<- c(error_files, paste0(validation_set_SIRIUS_compounds$validation_ID[i], '.ms'))
           message(paste0('Fingerprints missing: ', validation_set_SIRIUS_compounds$validation_ID[i]))
            print(e)
        }
    )
}


```

```{r}
orbitrap_files <- list.files(MS_folder_orbitrap)
orbitrap_files_again <- orbitrap_files[orbitrap_files %in% error_files]

qtof_files <- list.files(MS_folder_qtof)
qtof_files_again <- qtof_files[qtof_files %in% error_files]
```

After improving the `.ms`files with the data from MoNA, calculating the missing fingerprint features
```{r}
output_folder <- './SIRIUS_output_again'
MS2fp_SIRIUS(MS_folder_qtof, SIRIUS_folder, output_folder, 'qtof', qtof_files_again)
```

Combining the calculated fingerprint features with the toxicity information and writing the file that contains all the information needed for final evaluation of the models.
Also checking how well the true fingerprint features match with SIRIUS ones.
```{r}
together_csi_fingerid <- merge(csi_fingerid[, c(2, 14:640)], csi_fingerid_neg[, c(2, 14:120)], by='absoluteIndex')
together_csi_fingerid <- together_csi_fingerid[order(as.numeric(as.character(together_csi_fingerid$absoluteIndex))), ]
together_csi_fingerid$absoluteIndex <- sub('^', 'RelIdx_', together_csi_fingerid$absoluteIndex)
#write.table(together_csi_fingerid, 'calculated_validation_fps_SIRIUS.tsv', sep='\t', col.names=T, row.names=F, quote=F)
calculated_validation_fps_SIRIUS <- as.data.frame(together_csi_fingerid)
rownames(calculated_validation_fps_SIRIUS) <- calculated_validation_fps_SIRIUS[,1]
calculated_validation_fps_SIRIUS[, 1] <- NULL
calculated_validation_fps_SIRIUS_T <- transpose(calculated_validation_fps_SIRIUS)
colnames(calculated_validation_fps_SIRIUS_T) <- rownames(calculated_validation_fps_SIRIUS)
calculated_validation_fps_SIRIUS_T$validation_ID <- colnames(calculated_validation_fps_SIRIUS)

real_life_test_set_ready <- merge(validation_set_SIRIUS_compounds, calculated_validation_fps_SIRIUS_T, by='validation_ID')
#write.table(real_life_test_set_ready, 'real_life_test_set_ready.tsv', sep='\t', col.names=T, row.names=F, quote=F)
real_life_test_set_ready <- fread('real_life_test_set_ready.tsv', sep='\t',header=T)
fps_for_check <- fread('validation_set_fps.tsv', sep='\t', header=T)

train09 <- as.data.frame(fread('train_filtered_09.tsv', header=T, sep='\t'))

real_life_test_set_ready09 <- real_life_test_set_ready %>% select(colnames(train09[, c(1, 14:ncol(train09))]))
fps_for_check09 <- fps_for_check %>% select(colnames(train09[, c(1, 14:ncol(train09))]))
fps_for_check09 <- fps_for_check09 %>% filter(SMILES %in% real_life_test_set_ready09$SMILES)

real_life_test_set_ready09 <- as.data.frame(real_life_test_set_ready09[order(SMILES),])
fps_for_check09 <- as.data.frame(fps_for_check09[order(SMILES),])

check_fps <- data.frame(matrix(ncol=ncol(real_life_test_set_ready09), nrow=nrow(real_life_test_set_ready09)))

for (i in 1:nrow(real_life_test_set_ready09)) {
  check_fps[i, 1] <- real_life_test_set_ready09[i, 1]
  for (j in 2:ncol(real_life_test_set_ready09)) {
    check_fps[i, j] <- real_life_test_set_ready09[i, j] - fps_for_check09[i, j]
  }
}

colnames(check_fps) <- colnames(real_life_test_set_ready09)

melted_check_fps <- melt(check_fps)

ggplot(melted_check_fps, aes(x=variable, y=SMILES, fill=value)) + 
  geom_tile() +
  scale_fill_gradient2(low = '#4b2991', high = '#ea4f88', mid = '#edd9a3', 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="fps_SIRIUS - fps_calculated") +
  #theme_signature() +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank())


```
