---
title: "Mass spectrometric data preprocessing I"
output: html_notebook
---
```{r}
library(dplyr)
library(data.table)
library(stringr)
```

Reading all the files that are in MassBank (https://zenodo.org/record/7148841#.ZGEK_3ZBxpI) and merging the information into one file.
Each row in this combined file has the following information:
1. MassBank file name - for later searching (for generating the files for SIRIUS);
2. compound's SMILES - for mapping the compounds that have mass spectra with the compounds in the dataset obtained while running the code `Tox21_preprocessing_II.ipynb`;
3. instrument type (instrument used for measuring the MS) - for selecting only the HRMS data
```{r}
list_dirs_massbank <- list.dirs('./MassBank-data-2022.06a/', full.names=T)
mass_data <- data.frame(matrix(nrow=0, ncol=3))
colnames(mass_data) <- c('file_name', 'SMILES', 'instrument_type')

index = 1
for(i in 2:length(list_dirs_massbank)) {
  list_files <- list.files(list_dirs_massbank[i], full.names=T)
  for (file in list_files) {
    line_nr1 = grep('CH$SMILES: ', readLines(file), fixed=T)
    smiles_line = readLines(file)[line_nr1]
    SMILES = str_split(smiles_line, ' ')[[1]][2]
    
    line_nr2 = grep('AC$INSTRUMENT_TYPE: ', readLines(file), fixed=T)
    instrument_line = readLines(file)[line_nr2]
    instrument = str_split(instrument_line, ' ')[[1]][2]
    
    mass_data[index, 'file_name'] <- basename(file)
    mass_data[index, 'SMILES'] <- SMILES
    mass_data[index, 'instrument_type'] <- instrument
    index = index + 1
  }
}  

#write.table(mass_data, 'mass_data_info.tsv', row.names=F, col.names=T, sep='\t', quote=F)

# Filtering out some of the files based on the instrument type - not suitable for the final application (not high-resolution mass spectra)
mass_data_filtered <- mass_data[grep('ESI', mass_data$instrument_type), ]
mass_data_filtered <- mass_data_filtered[-grep('QQ', mass_data_filtered$instrument_type), ]

#write.table(mass_data_filtered, 'mass_data_info_filtered.tsv', row.names=F, col.names=T, sep='\t', quote=F)

mass_data_filtered <- mass_data_filtered %>% filter(instrument_type != 'LC-ESI-Q')
#write.table(mass_data_filtered, 'mass_data_to_standardisation.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

