---
title: "SMILES2fp"
author: "Ida Rahu"
date: "2022-10-12"
output: html_document
---
Code for calculating the true fingerprint features (that correspond to the molecular properties computed by SIRIUS+CSI:FingerID) for all the compounds in the Tox21 cleaned dataset.
```{r}
options(java.parameters = "- Xmx1024m")
library(rcdk)
library(rJava)
library(ChemmineR)
library(ChemmineOB)
library(dplyr)
library(fingerprint)
library(data.table)
library(tidyr)
library(reshape2)
library(stringr)

```

```{r}
tox_names <- c('nr.ahr', 'nr.ar.lbd', 'nr.ar', 'nr.aromatase', 'nr.er.lbd', 'nr.er', 'nr.ppar.gamma', 'sr.are', 'sr.atad5', 'sr.hse', 'sr.mmp', 'sr.p53')
```

```{r}
con <-file('feature_examples.txt')
lines <- readLines(con)
close(con)
slines <- strsplit(lines, '\t')
colCount <- max(unlist(lapply(slines, length)))

sirius_features <- data.frame(matrix(nrow=0, ncol=colCount))

for (i in 1:length(slines)) {
  line <- slines[[i]]
  for (j in 1:length(line)) {
    sirius_features[i, j] <- line[j]
  }
}

colnames(sirius_features)[1] <- 'unique_name'
colnames(sirius_features)[3] <- 'smarts_list'

#write.table(sirius_features, "sirius_features.tsv", row.names=F, col.names=T, sep='\t', quote=F)
```

Function for reading in the SMARTS patterns (from relevant files) that correspond to the SIRIUS+FingerID fingerprint features 
```{r}
pattern_file_reader <- function(file_name, split_pattern) {
  con <- file(file_name)
  lines <- readLines(con)
  close(con)
  slines <- strsplit(lines, split_pattern)
  colCount <- max(unlist(lapply(slines, length)))

  patterns <- data.frame(matrix(nrow=0, ncol=colCount))

  for (i in 1:length(slines)) {
    line <- slines[[i]]
    for (j in 1:length(line)) {
      patterns[i, j] <- line[j]
    }
  }
  return(patterns)
}
```

Reading in the cleaned Tox21 dataset (obtained from `Tox21_preprocessing_II.ipynb`)
```{r}
final_data <- fread('final_data_ready.tsv', header=T, sep='\t')
```

```{r}
# Fingerprints that cover all SIRIUS fingerprints (files downloaded from: https://github.com/boecker-lab/sirius)
OpenBabelFP3_names <- paste0('RelIdx_', c(0:54))
OpenBabelFP3_SMARTS <- unlist(pattern_file_reader('OpenBabel_FP3_patterns.txt', '\t')[1], use.names=F)

CDKsubstructure_names <- paste0('RelIdx_', c(55:361))
MACCS_names <- paste0('RelIdx_', c(362:527))
PubChem_names <- paste0('RelIdx_', c(528:1408))
KlekotaRoth_names <- paste0('RelIdx_', c(1409:6268))

ECFP6_names <- paste0('RelIdx_', c(6269:8178))
ECFP6_hashes <- unlist(read.table('ecfp_fp_hashes.txt', header=F), use.names=F)

custommadeSMARTS_names <- paste0('RelIdx_', c(8179:8461))
custommade_SMARTS <- unlist(pattern_file_reader('biosmarts_aka_custom_made_fps.txt', '\n'), use.names=F)

ringsystems_names <- paste0('RelIdx_', c(8462:8924))
ringsystems_SMARTS <- unlist(pattern_file_reader('ringsystem_fps.txt', '\n'), use.names=F)

```

Calculating the fingerprint features
```{r}
no_columns = ncol(final_data) + length(OpenBabelFP3_names) + length(CDKsubstructure_names) +
             length(MACCS_names) + length(PubChem_names) + length(KlekotaRoth_names) +
             length(ECFP6_names) + length(custommadeSMARTS_names) + length(ringsystems_names)


final_fp_data <- data.frame(matrix(nrow=nrow(final_data), ncol=no_columns))

colnames(final_fp_data) <- c(colnames(final_data), OpenBabelFP3_names, CDKsubstructure_names, MACCS_names, 
                             PubChem_names, KlekotaRoth_names, ECFP6_names, custommadeSMARTS_names,
                             ringsystems_names) 
final_fp_data[, c(1:ncol(final_data))] <- final_data

fp_index <- data.frame(matrix(nrow=8, ncol=3))
colnames(fp_index) <- c('fingerprint', 'start_index', 'end_index')
fp_index$fingerprint <- c('FP3', 'substructure', 'maccs', 'pubchem', 'kr', 'ecfp6', 'custom', 'ring')
fp_index$start_index <- c(ncol(final_data)+1, ncol(final_data)+length(OpenBabelFP3_names)+1, 
                          ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+1,
                          ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+1,
                          ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+1,
                          ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names)+1,
                          ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names)+length(ECFP6_names)+1,
                          ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names)+length(ECFP6_names)+length(custommadeSMARTS_names)+1)

fp_index$end_index <- c(ncol(final_data)+length(OpenBabelFP3_names), 
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names),
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names),
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names),
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names),
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names)+length(ECFP6_names),
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names)+length(ECFP6_names)+length(custommadeSMARTS_names),
                        ncol(final_data)+length(OpenBabelFP3_names)+length(CDKsubstructure_names)+length(MACCS_names)+length(PubChem_names)+length(KlekotaRoth_names)+length(ECFP6_names)+length(custommadeSMARTS_names)+length(ringsystems_names))

i = 1
for (SMILES in final_data$SMILES) {
  mol <- parse.smiles(SMILES)[[1]]
  if(!is.null(mol)){
    openbabel_fingerprints <- get.fingerprint(mol, type='substructure',
                                              substructure.pattern=OpenBabelFP3_SMARTS)
    final_fp_data[i, c(fp_index$start_index[1]:fp_index$end_index[1])] <- strsplit(as.character(openbabel_fingerprints), "")[[1]]
    
    substr_fingerprints <- get.fingerprint(mol, type='substructure')
    final_fp_data[i, c(fp_index$start_index[2]:fp_index$end_index[2])] <- strsplit(as.character(substr_fingerprints), "")[[1]]
      
    maccs_fingerprints <- get.fingerprint(mol, type='maccs')
    final_fp_data[i, c(fp_index$start_index[3]:fp_index$end_index[3])] <- strsplit(as.character(maccs_fingerprints), "")[[1]]

    pubchem_fingerprints <- get.fingerprint(mol, type='pubchem')
    final_fp_data[i, c(fp_index$start_index[4]:fp_index$end_index[4])] <- strsplit(as.character(pubchem_fingerprints), "")[[1]]

    kr_fingerprints <- get.fingerprint(mol, type='kr')
    final_fp_data[i, c(fp_index$start_index[5]:fp_index$end_index[5])] <- strsplit(as.character(kr_fingerprints), "")[[1]]

    ecfp_fingerprints <- get.fingerprint(mol, type='circular', circular.type='ECFP6', fp.mode='count')
    for (idx in 1:length(ecfp_fingerprints@features)) {
      hash <- strsplit(as.character(ecfp_fingerprints@features[[idx]]), ':')[[1]][1]
      if (hash %in% ECFP6_hashes) {
        right_hash <- which(hash == ECFP6_hashes)
        column_name <- ECFP6_names[right_hash]
        final_fp_data[i, column_name] <- 1
      }
    }

    custommade_fingerprints <- get.fingerprint(mol, type="substructure",
                                               substructure.pattern=custommade_SMARTS)
    final_fp_data[i, c(fp_index$start_index[7]:fp_index$end_index[7])] <- strsplit(as.character(custommade_fingerprints), "")[[1]]

    ring_fingerprints <- get.fingerprint(mol, type="substructure",
                                         substructure.pattern=ringsystems_SMARTS)
    final_fp_data[i, c(fp_index$start_index[8]:fp_index$end_index[8])] <- strsplit(as.character(ring_fingerprints), "")[[1]]
   
  }
  else {
    wrong_smile <- as.data.frame(SMILES)
    error_comp <- error_comp %>% rbind(wrong_smile)
    print(SMILES)
  }
  gc()
  i = i + 1
}

#write.table(final_fp_data, "final_data_with_fingerprints2.tsv", row.names=F, col.names=T, sep='\t', quote=F)
```

```{r}
final_fp_data2 <- final_fp_data

final_fp_data2 <- final_fp_data2 %>% mutate_at(ECFP6_names, ~replace_na(as.numeric(.), 0))
final_fp_data2 <- final_fp_data2 %>% mutate_at(c(3:ncol(final_fp_data2)), as.numeric)

#write.table(final_fp_data2, "final_data_with_fingerprints_withoutNA.tsv", row.names=F, col.names=T, sep='\t', quote=F)
```
Generating the names for features so calculated fingerprint features would match with SIRIUS+CSI:FingerID absolute index.
```{r}
positive_idxs <- pattern_file_reader('csi_fingerid.tsv', '\t')
colnames(positive_idxs) <- positive_idxs[1,]
positive_idxs <- positive_idxs[-1, ] 

negative_idxs <- pattern_file_reader('csi_fingerid_neg.tsv', '\t')
colnames(negative_idxs) <- negative_idxs[1,]
negative_idxs <- negative_idxs[-1, ] 

together_idx <- merge(positive_idxs, negative_idxs, by='absoluteIndex')
together_idx <- together_idx[order(as.numeric(as.character(together_idx$absoluteIndex))), ]
positive_idxs$absoluteIndex <- sub('^', 'RelIdx_', positive_idxs$absoluteIndex)
negative_idxs$absoluteIndex <- sub('^', 'RelIdx_', negative_idxs$absoluteIndex)
together_idx$absoluteIndex <- sub('^', 'RelIdx_', together_idx$absoluteIndex)

```

In order to build classification models that are applicable regardless of which ionisation mode (positive or negative) was used during the HRMS analysis selecting the overlapping features (features that are always outputted by SIRIUS+FingerID)
```{r}
positive_mode_data <- final_fp_data2 %>% dplyr::select(c('SMILES', tox_names, positive_idxs$absoluteIndex))
write.table(positive_mode_data, 'positive_mode_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)

negative_mode_data <- final_fp_data2 %>% dplyr::select(c('SMILES', tox_names, negative_idxs$absoluteIndex))
write.table(negative_mode_data, 'negative_mode_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)

general_mode_data <- final_fp_data2 %>% dplyr::select(c('SMILES', tox_names, together_idx$absoluteIndex))
write.table(general_mode_data, 'general_mode_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)

general_mode_data <- fread('general_mode_data.tsv', header=T, sep='\t')
```

Removing the features that correspond to the structural patterns that do not exist in any of the compounds.
```{r}
general_col_sums <- colSums(general_mode_data %>% select(c(14:ncol(general_mode_data))))

general_mode_data_without_all0 <- general_mode_data %>% select(-names(which(general_col_sums==0)))
write.table(general_mode_data_without_all0, 'general_mode_data_without_all0.tsv', row.names=F, col.names=T, sep='\t', quote=F)

n0 = which(general_col_sums==0)

# Checking if there are features that correspond to the structural patterns that exist in all of the compounds
#which(general_col_sums==7483)
```

Visualising the correlation between the features
```{r}
fingerprint_features_general <- general_mode_data_without_all0 %>% select(c(1, 14:ncol(general_mode_data_without_all0)))

cor_matrix <- cor(fingerprint_features_general %>% select(c(2:ncol(fingerprint_features_general))))

melted_cor_matrix <- melt(cor_matrix)
```

```{r}
library(ggplot2)
theme_signature <- function() {
  theme_minimal(base_size = 14, base_family = "Rubik") +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(size = .13, color = '#E1E1E1'),
      axis.text = element_text(color = "#929497"),
      axis.ticks.x = element_line(size = 0.5, color = '#E1E1E1'),
      axis.ticks.y = element_line(size = 0.5, color = '#E1E1E1'),
      axis.title = element_text(color = "white"),
      axis.title.y = element_text(hjust = 1, margin = margin(0, 6, 0, 15, "pt")),
      axis.title.x = element_text(hjust = 0, margin = margin(6, 0, 15, 0, "pt")),
      plot.subtitle = element_text(color = "white", size= 11),
      plot.title = element_text(color = "white", size= 20),
      plot.title.position = "plot",
      plot.caption = element_text(hjust = 0, color = "white"),
      plot.caption.position = "plot",
      plot.margin = margin(.5,.5,.5,.5,"cm"),
      plot.background = element_rect(fill = "#343434"),
      legend.text = element_text(color = "#E1E1E1"),
      legend.title = element_text(color = "#E1E1E1"),
      strip.text = element_text(color = "white")) 
}
```

```{r}
library(viridis)
ggplot(melted_cor_matrix, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  scale_fill_gradient2(low = '#4b2991', high = '#edd9a3', mid = '#ea4f88', 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="correlation") +
  theme_signature() +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank())

ggsave('correlation.png', dpi=720)
```
