---
title: "Training and intermediate testing sets"
author: "Ida Rahu"
date: "2022-10-12"
output: html_document
---
Code for removing the highly correlated fingerprint features and features with zero or near-zero variance, and preparing the data for training the models (splitting data into training and intermediate testing sets).
```{r}
library(dplyr)
library(data.table)
library(tidyr)
library(reshape2)
library(stringr)
library(ggplot2)
library(viridis)
library(caret)
library(anticlust)
```

```{r}
tox_names <- c('nr.ahr', 'nr.ar.lbd', 'nr.ar', 'nr.aromatase', 'nr.er.lbd', 'nr.er', 'nr.ppar.gamma', 'sr.are', 'sr.atad5', 'sr.hse', 'sr.mmp', 'sr.p53')
```

```{r}

theme_signature <- function() {
  theme_minimal(base_size = 14, base_family = "Rubik") +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(size = .13, color = '#E1E1E1'),
      axis.text = element_text(color = "#929497"),
      axis.ticks.x = element_line(size = 0.5, color = '#E1E1E1'),
      axis.ticks.y = element_line(size = 0.5, color = '#E1E1E1'),
      axis.title = element_text(color = "white"),
      axis.title.y = element_text(hjust = 1, margin = margin(0, 6, 0, 15, "pt")),
      axis.title.x = element_text(hjust = 0, margin = margin(6, 0, 15, 0, "pt")),
      plot.subtitle = element_text(color = "white", size= 11),
      plot.title = element_text(color = "white", size= 20),
      plot.title.position = "plot",
      plot.caption = element_text(hjust = 0, color = "white"),
      plot.caption.position = "plot",
      plot.margin = margin(.5,.5,.5,.5,"cm"),
      plot.background = element_rect(fill = "#343434"),
      legend.text = element_text(color = "#E1E1E1"),
      legend.title = element_text(color = "#E1E1E1"),
      strip.text = element_text(color = "white")) 
}
```

Reading in the information about compounds that form the training and intermediate testing sets (generated in `Split_data_real_life_test_set.ipynb`)
```{r}
train_test <- fread("train_test.tsv", header=T, sep='\t')
```


```{r}
# Removing the fingerprint features with zero and near-zero variance
near_zero_var_fp <- nearZeroVar(train_test[,-c(1:13)], saveMetrics=T, names=T)
near_zero_var_fp$fp_features <- rownames(near_zero_var_fp)

near_zero_var_fp_filter <- near_zero_var_fp %>% filter(nzv == TRUE)

filtered_fp <- as.data.frame(train_test)[, -which(names(train_test) %in% near_zero_var_fp_filter$fp_features)]

# Removing the highly correlated fingerprint features
# Using three different cutoff values: 0.7, 0.8 and 0.9
correlation_matrix <- cor(filtered_fp[,-c(1:13)])
correlation_fp_names_07 <- findCorrelation(correlation_matrix, cutoff=0.7, names=T) 
correlation_fp_names_08 <- findCorrelation(correlation_matrix, cutoff=0.8, names=T)
correlation_fp_names_09 <- findCorrelation(correlation_matrix, cutoff=0.9, names=T)

filtered_fp_07 <- as.data.frame(filtered_fp)[, -which(names(filtered_fp) %in% correlation_fp_names_07)]
filtered_fp_08 <- as.data.frame(filtered_fp)[, -which(names(filtered_fp) %in% correlation_fp_names_08)]
filtered_fp_09 <- as.data.frame(filtered_fp)[, -which(names(filtered_fp) %in% correlation_fp_names_09)]

#write.table(filtered_fp_07, "filtered_train_test_07.tsv", row.names=F, col.names=T, sep='\t', quote=F)
#write.table(filtered_fp_08, "filtered_train_test_08.tsv", row.names=F, col.names=T, sep='\t', quote=F)
#write.table(filtered_fp_09, "filtered_train_test_09.tsv", row.names=F, col.names=T, sep='\t', quote=F)
```

```{r}
filtered_fp07 <- fread("filtered_train_test_07.tsv", sep='\t', header=T)
filtered_fp08 <- fread("filtered_train_test_08.tsv", sep='\t', header=T)
filtered_fp09 <- fread("filtered_train_test_09.tsv", sep='\t', header=T)
filtered_fp07[is.na(filtered_fp07)] <- 99999
filtered_fp08[is.na(filtered_fp08)] <- 99999
filtered_fp09[is.na(filtered_fp09)] <- 99999
```

Using anticulstering algorithm to generate the training and intermediate testing sets.
```{r}
equal <- anticlustering(
  filtered_fp07[, c('nr.ahr', 'nr.ar.lbd', 'nr.ar', 'nr.aromatase', 
                    'nr.er.lbd', 'nr.er', 'nr.ppar.gamma', 'sr.are',
                    'sr.atad5', 'sr.hse', 'sr.mmp', 'sr.p53')], 
  K = 5, 
  objective="diversity",
  standardize=F 
)
```

```{r}
filtered_fp07$groups <- equal
filtered_fp08$groups <- equal
filtered_fp09$groups <- equal

test_filtered07 <- filtered_fp07 %>% filter(groups == 5) %>% dplyr::select(-groups)
train_filtered07 <- filtered_fp07 %>% filter(groups < 5) %>% dplyr::select(-groups)

test_filtered08 <- filtered_fp08 %>% filter(groups == 5) %>% dplyr::select(-groups)
train_filtered08 <- filtered_fp08 %>% filter(groups < 5) %>% dplyr::select(-groups)

test_filtered09 <- filtered_fp09 %>% filter(groups == 5) %>% dplyr::select(-groups)
train_filtered09 <- filtered_fp09 %>% filter(groups < 5) %>% dplyr::select(-groups)

train_summary07 <- data.frame(matrix(nrow=0, ncol=4))
test_summary07 <- data.frame(matrix(nrow=0, ncol=4))

train_summary08 <- data.frame(matrix(nrow=0, ncol=4))
test_summary08 <- data.frame(matrix(nrow=0, ncol=4))

train_summary09 <- data.frame(matrix(nrow=0, ncol=4))
test_summary09 <- data.frame(matrix(nrow=0, ncol=4))

for (i in 2:13) {
  dat1 <- cbind(tox=tox_names[i-1], t(table(train_filtered07[, ..i])))
  dat2 <- cbind(tox=tox_names[i-1], t(table(test_filtered07[, ..i])))
  train_summary07 <- rbind(train_summary07, dat1)
  test_summary07 <- rbind(test_summary07, dat2)
  
  dat3 <- cbind(tox=tox_names[i-1], t(table(train_filtered08[, ..i])))
  dat4 <- cbind(tox=tox_names[i-1], t(table(test_filtered08[, ..i])))
  train_summary08 <- rbind(train_summary08, dat3)
  test_summary08 <- rbind(test_summary08, dat4)
  
  dat5 <- cbind(tox=tox_names[i-1], t(table(train_filtered09[, ..i])))
  dat6 <- cbind(tox=tox_names[i-1], t(table(test_filtered09[, ..i])))
  train_summary09 <- rbind(train_summary09, dat5)
  test_summary09 <- rbind(test_summary09, dat6)
}

colnames(train_summary07) <- c('tox_name', 'non_active_train', 'active_train', 'missing_train')
colnames(test_summary07) <- c('tox_name', 'non_active_test', 'active_test', 'missing_test')

colnames(train_summary08) <- c('tox_name', 'non_active_train', 'active_train', 'missing_train')
colnames(test_summary08) <- c('tox_name', 'non_active_test', 'active_test', 'missing_test')

colnames(train_summary09) <- c('tox_name', 'non_active_train', 'active_train', 'missing_train')
colnames(test_summary09) <- c('tox_name', 'non_active_test', 'active_test', 'missing_test')

train_summary <- train_summary07
test_summary <- test_summary07

train_summary[, 2:ncol(train_summary)] <- apply(train_summary[, 2:ncol(train_summary)], 2,           
                    function(x) as.numeric(as.character(x)))

test_summary[, 2:ncol(test_summary)] <- apply(test_summary[, 2:ncol(test_summary)], 2,           
                    function(x) as.numeric(as.character(x)))

train_summary <- train_summary %>% mutate(usable_train = non_active_train + active_train)
train_summary <- train_summary %>% mutate(active_pct_train = active_train/usable_train)

test_summary <- test_summary %>% mutate(usable_test = non_active_test + active_test)
test_summary <- test_summary %>% mutate(active_pct_test = active_test/usable_test)

tog <- merge(train_summary, test_summary, by='tox_name')
tog <- tog %>% mutate(usable_test_train_ratio = usable_test/usable_train)
tog <- tog %>% mutate(non_active_test_train_ratio = non_active_test/non_active_train)
tog <- tog %>% mutate(active_test_train_ratio = active_test/active_train)

```

Visualising the dataset.
```{r}
melt_tog <- melt(tog)
melt_active <- melt_tog %>% filter(variable %in% c('active_pct_train', 'active_pct_test'))

ggplot(melt_active, aes(x=tox_name, y=value, fill=as.factor(variable), color=as.factor(variable))) +
  geom_bar(stat='identity', alpha=0.4, position='dodge') +
  scale_fill_manual('', values=c('#B1339E', '#08f26e'), label=c('model train', 'model test')) +
  scale_color_manual('', values=c('#B1339E', '#08f26e'), label=c('model train', 'model test')) +
  scale_y_continuous(labels=scales::percent) +
  xlab('assay') +
  ylab('percentage of active compounds (%)') +
  theme_signature() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size=8))

ggsave('active_pct_per_assay_model_test_train.png', dpi=400)
```

```{r}
options(warn=-1)
ggplot(tog, aes(x=tox_name)) +
  geom_point(aes(y=usable_test_train_ratio), color='#e1e1e1', size=4, alpha=0.5) +
  geom_text(aes(x=0.5, y=0.25), label='all', color='#e1e1e1') +
  geom_point(aes(y=non_active_test_train_ratio), color='#B1339E', size=4, alpha=0.5) +
  geom_text(aes(x=1, y=0.3), label='non-active', color='#B1339E') +
  geom_point(aes(y=active_test_train_ratio), color='#08f26e', size=4, alpha=0.5) +
  geom_text(aes(x=1, y=0.2), label='active', color='#08f26e') +
  xlab('assay') +
  ylab(expression(paste(frac(italic(n)[model_test], italic(n)[model_train])))) +
  expand_limits(x=-0.25, y=0.015) +
  theme_signature() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size=8))

ggsave('ratios_model_test_train.png', dpi=400)
```

Writing the obtained training and intermediate testing sets into relevant files.
```{r}
write.table(test_filtered07, "test_filtered_07.tsv", row.names=F, col.names=T, sep='\t', quote=F)
write.table(train_filtered07, "train_filtered_07.tsv", row.names=F, col.names=T, sep='\t', quote=F)

write.table(test_filtered08, "test_filtered_08.tsv", row.names=F, col.names=T, sep='\t', quote=F)
write.table(train_filtered08, "train_filtered_08.tsv", row.names=F, col.names=T, sep='\t', quote=F)

write.table(test_filtered09, "test_filtered_09.tsv", row.names=F, col.names=T, sep='\t', quote=F)
write.table(train_filtered09, "train_filtered_09.tsv", row.names=F, col.names=T, sep='\t', quote=F)
```

```{r}
# Writing training and intermediate test sets into separate files based on the assay
for (i in 2:13) {
  name <- colnames(test_filtered07)[i]
  data07 <- test_filtered07 %>% dplyr::select(c(i, 14:ncol(test_filtered07)))
  write.table(data07, paste0('./data_for_models/data_07/test/test_', name, '.tsv'), row.names=F, col.names=T, sep='\t', quote=F)
  tdata07 <- train_filtered07 %>% dplyr::select(c(i, 14:ncol(train_filtered07)))
  write.table(tdata07, paste0('./data_for_models/data_07/train/', name, '.tsv'), row.names=F, col.names=T, sep='\t', quote=F)
  
  data08 <- test_filtered08 %>% dplyr::select(c(i, 14:ncol(test_filtered08)))
  write.table(data08, paste0('./data_for_models/data_08/test/test_', name, '.tsv'), row.names=F, col.names=T, sep='\t', quote=F)
  tdata08 <- train_filtered08 %>% dplyr::select(c(i, 14:ncol(train_filtered08)))
  write.table(tdata08, paste0('./data_for_models/data_08/train/', name, '.tsv'), row.names=F, col.names=T, sep='\t', quote=F)
  
  data09 <- test_filtered09 %>% dplyr::select(c(i, 14:ncol(test_filtered09)))
  write.table(data09, paste0('./data_for_models/data_09/test/test_', name, '.tsv'), row.names=F, col.names=T, sep='\t', quote=F)
  tdata09 <- train_filtered09 %>% dplyr::select(c(i, 14:ncol(train_filtered09)))
  write.table(tdata09, paste0('./data_for_models/data_09/train/', name, '.tsv'), row.names=F, col.names=T, sep='\t', quote=F)
}
```

