---
title: "Tox21 data preprocessing I"
output: html_notebook
---

```{r}
library(dplyr)
library(data.table)
```

Merging all the input (.smiles; https://tripod.nih.gov/tox21/challenge/) files (every assay has a separate file)
```{r}
files_smiles <- list.files('./SMILES', pattern='*.smiles', full.names=T)

tox_names <- tools::file_path_sans_ext(lapply(files_smiles, basename))
tox_names <- gsub('-', '.', tox_names)

data_collected = data.frame(matrix(nrow=0, ncol=2))
colnames(data_collected) <- c('SMILES', 'ID')

for (i in 1:length(files_smiles)) {
  data <- fread(files_smiles[i], sep='\t')
  colnames(data) <- c('SMILES', 'ID', tox_names[i])
  data_collected <- merge(data_collected, data, by=c('SMILES', 'ID'), all=T)
}
#write.table(data_collected, 'data_collected.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

Finding the duplicated compounds
```{r}
duplicates <- subset(data_collected$SMILES, duplicated(data_collected$SMILES))
duplicates_un <- unique(duplicates)

duplicates_collected <- data.frame(matrix(nrow=0, ncol=ncol(data_collected)))
colnames(duplicates_collected) <- colnames(data_collected)

unique_collected <- data.frame(matrix(nrow=0, ncol=ncol(data_collected)))
colnames(unique_collected) <- colnames(data_collected)

for (i in 1:nrow(data_collected)) {
  if (data_collected$SMILES[i] %in% duplicates_un) {
    duplicates_collected <- rbind(duplicates_collected, data_collected[i, ])
  }
  else {
    unique_collected <- rbind(unique_collected, data_collected[i, ])
  }
}

#write.table(duplicates_collected, 'duplicates_collected.tsv', row.names=F, col.names=T, sep='\t', quote=F)
#write.table(unique_collected, 'unique_collected.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

Deduplicating the dataset

Collecting the duplicates and then comparing the activity information in each assay.

Based on that combining the compound's data in each assay using the following rules.

If the value for this assay:

1. is **missing** in each duplicated row > the value is `NA` in the final dataset;
2. is **zero** (inactive) or **one** (active) in each duplicated row > the value is `0` or `1` (respectively) in the final dataset;
3. is **one** in at least one of the duplicated rows (no matter what are the other values) > the value is `1` in the final dataset;
4. is **zero** in at least one of the duplicated rows and missing in the other rows > the value is `0` in the final dataset. 
```{r}
summarised_duplicates <- data.frame(matrix(nrow=0, ncol=ncol(data_collected)))
colnames(summarised_duplicates) <- colnames(data_collected)

changed_activity <- data.frame(matrix(nrow=0, ncol=2))
colnames(changed_activity) <- c('SMILES', 'tox_name')

index = 1
for (i in 1:length(duplicates_un)) {
  duplicates_to_summarise <- duplicates_collected[duplicates_collected$SMILES == duplicates_un[i], ]
  summarised_duplicates[i, c(1:2)] <- duplicates_to_summarise[1, c(1:2)]
  SMILES_name <- duplicates_to_summarise[1, 1]
  for(j in 1:length(tox_names)) {
    tox_end <- duplicates_to_summarise[, tox_names[j]]
    tox_end_without_NA <- tox_end[!is.na(tox_end)]
    if (length(tox_end_without_NA) > 0) {
      if (1 %in% tox_end_without_NA) {
        summarised_duplicates[i, tox_names[j]] <- 1
        if (0 %in% tox_end_without_NA) {
          changed_activity[index, 'SMILES'] <- SMILES_name
          changed_activity[index, 'tox_name'] <- tox_names[j]
          index <- index + 1 
        }
      }
      else {
        summarised_duplicates[i, tox_names[j]] <- 0
      }
    }
    else {
      summarised_duplicates[i, tox_names[j]] <- NA
    }
  }
}

final_data <- rbind(unique_collected, summarised_duplicates)

#write.table(final_data, 'final_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

Checking how many compounds in each assay are marked as active in the final dataset because of rule number 3.
```{r}
changed_activity_all <- data_collected %>% filter(SMILES %in% changed_activity$SMILES)
#write.table(changed_activity_all,'changed_activity_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)

changed_activity_sum <- changed_activity %>% group_by(tox_name) %>% tally()
chem_changed_act_sum <- changed_activity %>% group_by(SMILES) %>% tally()

data_summary <- data.frame(matrix(nrow=0, ncol=4))

for (i in 3:14) {
  dat <- cbind(tox=tox_names[i-2], t(table(final_data[, i], useNA='always')))
  data_summary <- rbind(data_summary, dat)
}
colnames(data_summary) <- c('tox_name', 'non_active_full', 'active_full', 'missing_full')

data_summary <- merge(data_summary, changed_activity_sum, by='tox_name')
colnames(data_summary)[5] <- 'changed_activity'

data_summary[, 2:ncol(data_summary)] <- apply(data_summary[, 2:ncol(data_summary)], 2,           
                    function(x) as.numeric(as.character(x)))
data_summary <- data_summary %>% mutate(pct_changed_act = changed_activity/active_full)
data_summary
```
